---
title: "OAuth2 Setup for Brightspace Data Sets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{OAuth2 Setup for Brightspace Data Sets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

Before you can use brightspaceR, you need to register an OAuth2 application in
your Brightspace instance and configure your R environment with the credentials.
This guide walks through every step.

## Step 1: Register an OAuth2 Application

You need administrator access (or help from an admin) to register an app.

1. Log in to your Brightspace instance (e.g., `https://myschool.brightspace.com`).
2. Navigate to **Admin Tools** > **Manage Extensibility**.
3. Select the **OAuth 2.0** tab.
4. Click **Register an app**.
5. Fill in the registration form:

| Field | Value |
|---|---|
| **Application Name** | `brightspaceR` (or any name you like -- this is shown to users on the consent page) |
| **Authentication Workflow** | **Authorization Code Grant** |
| **Redirect URI** | `https://localhost:1410/` |
| **Scope** | `datahub:dataexports:*` |
| **Access Token Lifetime** | Leave at default (3600 seconds is fine) |
| **Enable Refresh Tokens** | **Yes** (recommended) |
| **Prompt User for Consent** | Optional (Yes if you want users to see a consent screen) |

6. Accept the **Non-Commercial Developer Agreement** (or your institution's
   terms).
7. Click **Register**.

After registration you'll receive a **Client ID** and **Client Secret**. Save
these securely -- you'll need them in the next step.

### About the Redirect URI

Brightspace requires all redirect URIs to use HTTPS -- `http://` is not
accepted. brightspaceR uses `https://localhost:1410/` by default.

**How the flow works:** When you run `bs_auth()`, the package opens your
browser to the Brightspace authorization page. After you authorize, Brightspace
redirects to `https://localhost:1410/?code=...`. Since there's no local HTTPS
server listening, your browser will show a connection error -- that's expected.
You simply copy the full URL from your browser's address bar and paste it back
into R. The package extracts the authorization code from the URL and exchanges
it for an access token.

You can use a different redirect URI by setting the `BRIGHTSPACE_REDIRECT_URI`
environment variable or passing `redirect_uri` to `bs_auth()`. Just make sure
it matches what you registered in Brightspace exactly.

### About Scopes

Scopes control what API endpoints your application can access. Brightspace
scopes follow the pattern `<group>:<resource>:<action>`.

| Scope | Access |
|---|---|
| `datahub:dataexports:*` | Brightspace Data Sets (BDS) -- full and differential extracts |
| `datahub:adhocdataexports:*` | Advanced Data Sets (ADS) -- ad-hoc data exports |
| `core:*:*` | General API access (users, enrollments, etc. -- not needed for BDS) |

For brightspaceR you only need `datahub:dataexports:*`. If you also want
Advanced Data Sets, register with both scopes separated by a space:
`datahub:dataexports:* datahub:adhocdataexports:*`.

## Step 2: Configure Your R Environment

There are two ways to store your credentials: a **config file** (recommended
for projects) or **environment variables** (traditional approach). Both are
picked up automatically by `bs_auth()`.

### Option A: Config file (recommended)
{#config-file}

Create a `config.yml` file in your project root:

```yaml
default:
  brightspace:
    client_id: "your-client-id"
    client_secret: "your-client-secret"
    instance_url: "https://myschool.brightspace.com"
    redirect_uri: "https://localhost:1410/"
```

Or use `bs_config_set()` to create it interactively:
```{r}
bs_config_set(
  client_id = "your-client-id",
  client_secret = "your-client-secret",
  instance_url = "https://myschool.brightspace.com"
)
```

The config file supports environment-based profiles via the
[config](https://rstudio.github.io/config/) package. Set the
`R_CONFIG_ACTIVE` environment variable to switch profiles:

```yaml
default:
  brightspace:
    client_id: "dev-id"
    instance_url: "https://dev.brightspace.com"

production:
  inherits: default
  brightspace:
    client_id: "prod-id"
    instance_url: "https://myschool.brightspace.com"
```

> **Security note**: Make sure `config.yml` is in your `.gitignore` to avoid
> committing secrets to version control.

### Option B: Environment variables

You can pass credentials directly to `bs_auth()`, but it's more convenient (and
safer) to store them as environment variables. Add the following to your
`.Renviron` file:
```{r}
# Open .Renviron for editing
usethis::edit_r_environ()
```

Then add these lines:
```
BRIGHTSPACE_CLIENT_ID=your-client-id-here
BRIGHTSPACE_CLIENT_SECRET=your-client-secret-here
BRIGHTSPACE_INSTANCE_URL=https://myschool.brightspace.com
```

Restart R for the changes to take effect.

> **Security note**: Never commit `.Renviron` to version control. It should
> already be listed in `.gitignore` by default.

### Credential resolution order

`bs_auth()` resolves each credential in this order:

1. Explicit argument (e.g., `bs_auth(client_id = "...")`)
2. `config.yml` in the working directory (if present)
3. Environment variable (e.g., `BRIGHTSPACE_CLIENT_ID`)

## Step 3: Authenticate

```{r}
library(brightspaceR)

# Uses environment variables automatically
bs_auth()
```

This will:

1. Open your default browser to the Brightspace login/consent page.
2. You log in with your Brightspace credentials and authorize the app.
3. Brightspace redirects to `https://localhost:1410/?code=...&state=...`.
4. Your browser shows a connection error (because there's no local HTTPS
   server) -- **this is expected and normal**.
5. Copy the entire URL from your browser's address bar.
6. Paste it into the R console when prompted.
7. brightspaceR extracts the authorization code, exchanges it for tokens, and
   caches them to disk.

You should see:
```
v Authenticated with Brightspace at <https://myschool.brightspace.com>
```

### Token Caching and Refresh

Tokens are cached to disk automatically. On subsequent calls, `bs_auth()` will
reuse the cached token and refresh it if expired -- no browser interaction
needed.

To force re-authentication:
```{r}
bs_deauth()
bs_auth()
```

### Non-interactive Environments (Scheduled Scripts)

For scripts that run unattended (e.g., cron jobs, scheduled ETL pipelines), use
a refresh token obtained from a prior interactive session:
```{r}
bs_auth_refresh(
  refresh_token = Sys.getenv("BRIGHTSPACE_REFRESH_TOKEN")
)
```

This exchanges the refresh token for a new access token without any browser
interaction. Store the refresh token as an environment variable or in a
secure secrets manager.

## Step 4: Verify It Works

```{r}
# Check authentication status
bs_has_token()
#> [1] TRUE

# List available datasets
datasets <- bs_list_datasets()
datasets
#> # A tibble: 67 x 5
#>    schema_id plugin_id name               description             created_date
#>    <chr>     <chr>     <chr>              <chr>                   <chr>
#>  1 abc123    def456    Users              User demographics ...   2024-01-01...
#>  2 ghi789    jkl012    User Enrollments   Enrollment records ...  2024-01-01...
#> ...
```

## Troubleshooting

### "No client ID found"

Make sure your environment variables are set. Check with:
```{r}
Sys.getenv("BRIGHTSPACE_CLIENT_ID")
```

If empty, re-check your `.Renviron` file and restart R.

### Browser doesn't open

If you're in an environment without a browser (e.g., RStudio Server), the
authorization URL is printed to the console. Copy it into a browser on any
machine, authorize, then copy the redirect URL back.

### "No authorization code found in the redirect URL"

Make sure you copy the **entire** URL from the address bar, including the
`?code=...&state=...` query parameters. It should start with
`https://localhost:1410/`.

### 403 Forbidden errors

Your OAuth2 app may not have the correct scope, or the user account you
authenticated with may not have permission to access Data Hub. Check:

- The registered scope includes `datahub:dataexports:*`.
- Your Brightspace user role has the **Data Hub** permissions enabled
  (typically requires an admin or a role with "Can Access Data Hub" permission).

### Token expired and won't refresh

Clear the cached token and re-authenticate:
```{r}
bs_deauth()
bs_auth()
```

## References

- [Brightspace OAuth 2.0 Documentation](https://docs.valence.desire2learn.com/basic/oauth2.html)
- [Getting Started with OAuth 2.0](https://community.d2l.com/brightspace/kb/articles/21863-how-to-get-started-with-oauth-2-0)
- [Getting Started with Data Hub APIs: Brightspace Data Sets](https://community.d2l.com/brightspace/kb/articles/1130-how-to-get-started-with-data-hubs-apis-brightspace-data-sets)
- [BDS Headless Client Example (D2L GitHub)](https://github.com/Brightspace/bds-headless-client-example)
